---
title: "Tidying Data in R"
author: "Kelsey Aho"
date: "August 15, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read in libraries. The tidyverse comprises of: ggplot2, dplyr, tidyr, tibble, purr, readr
```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
```

Read in data from [Mike Byerly. 2016. Alaska commercial salmon catches by management region (1886- 1997). Gulf of Alaska Data Portal. df35b.304.2.](https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1). 
```{r}
catch_df <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"),
                    stringsAsFactors = FALSE)
head(catch_df)
```

Both **tidyr** and **diplyr** use the pipe operator %>%
Pipe operator: CONTROL + SHIFT + m >> %>% 

Below, I select the columns that I will be using in a later analysis: 
```{r}
catch_df <- catch_df %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_df)
```


- Each line using the **pipe** operator must end with the pipe if there is a function on the next line.
Below, I remove the columns that I will not be using in a later analysis: 
```{r}
#catch_df <- catch_df %>%
#  select(-All, -notesRegCode)

#head(catch_df)
```


**Gather** will change the table from wide to long format. Region and year will be left as is. 
**Rename** will replace the catch column name with catch_thousands.
```{r}
catch_df <- catch_df %>% 
  gather(key= "species", value="catch", -Region, -Year) %>% 
  rename(catch_thousands = catch)

```

Reverse of **gather** is **spread**

The catch_thousands column are characters. **As** is used to change the type to integer.
```{r}
catch_integers <- as.integer(catch_df$catch_thousands)
i <- which(is.na(catch_integers) == T)
catch_df$catch_thousands[i]
```

**Mutate** will be used to first replace the *I* value with *1*.
Second, **Mutate** creates a new column with the actual number of catch.
```{r}
catch_df <- catch_df %>% 
  mutate(catch_thousands = ifelse(catch_thousands == "I", 1, catch_thousands)) %>% 
  mutate(catch_thousands = as.integer(catch_thousands)) %>% 
  mutate(catch = catch_thousands * 1000)

head(catch_df)
```






